---
title: "README.md"
author: "Soumya R Das"
date: "May 1, 2016"
output: md_document
---

## Getting and Cleaning Data Course Project

### Requirements for the Project:
> Create one R script named run_analysis.R that does the following:
>
  * The submitted data set is tidy.
  * Merges the training and the test sets to create one data set.
  * Extracts only the measurements on the mean and standard deviation for each     measurement.
  * Uses descriptive activity names to name the activities in the data set
  * Appropriately labels the data set with descriptive variable names.
  * From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
  
### Additional submission requirements for the Project are: 
> 
  * The submitted data set is tidy.
  * The Github repo contains the required scripts.
  * GitHub contains a code book that modifies and updates the available codebooks with the data to indicate all the variables and summaries calculated, along with units, and any other relevant information.
  * The README that explains the analysis files is clear and understandable.
  * The work submitted for this project is the work of the student who submitted it.
  
### Overall Process:  
The below process was followed to arrive at the two tidy data sets required for the project.

#### 1. Merge the training and the test sets to create one data set.
I am using the dplyr package to select, rename and mutate data tables in this R script, so I load the package first. I also load the activity labels and the features data into data tables.

```{r eval=FALSE}
# Read the activity labels and features data sets that are common to both train and test data
activity_labels <- read.table("./activity_labels.txt")
features <- read.table("./features.txt")
```

I also read the first set of test data from the files into data tables X_test, Y_test and subject_test 

```{r eval=FALSE}
# Read test data
X_test <- read.table("./test/X_test.txt")
Y_test <- read.table("./test/Y_test.txt")
subject_test <- read.table("./test/subject_test.txt")
```

Similar code can be written to read the second set of training data from the files into data tables X_train, Y_train and subject_train

```{r eval=FALSE}
# Read training data
X_train <- read.table("./train/X_train.txt")
Y_train <- read.table("./train/Y_train.txt")
subject_train <- read.table("./train/subject_train.txt")
```

After reading test and train data, they are merged using the rbind() function. The order of rows have to be maintained in all the resultant data sets X_test_train, Y_test_train and subject_test_train - first the test data and then the training data
```{r, eval=FALSE}
X_test_train <- rbind(X_test,X_train)
Y_test_train <- rbind(Y_test,Y_train)
subject_test_train <- rbind(subject_test, subject_train)
```

#### 2. Extract only the measurements on the mean and standard deviation for each     measurement.
I have already read the set of features into the features data table previously. From that set, I filter out the ones that contain "mean" or "std" (assuming std stands for standard deviation) - using the grep() function for searching and the filter() verb of dplyr for filtering out the data set.

NOTE: There are some features(or measurements) which contain something like meanFreq which I am excluding for my analysis. The resulting feature set for analysis is stored in a data table called sub_features
```{r eval=FALSE}
# Extract only the measurements on the mean and standard deviation for each measurement.
sub_features <- filter(features, grepl("mean|std",V2), !grepl("meanFreq",V2))
```

I then use the sub_features data set to modify the X_test_train data set to only contain variables which have mean or std measurements as given below.
```{r eval=FALSE}
# Use only the previously obtained subset of variables/measurements in the final subset
X_test_train <- X_test_train[,sub_features$V2]
```

#### 3. Use descriptive activity names to name the activities in the data set
Before the final data set is built, I rename the variables in the X_test_train by using the row values in the V2 column of sub_features. These are basically the names of the measurements(mean and std) that we are considering for this analysis.

I also rename some variables to make more meaningful names like given below. For example, V1 column of activity_labels is named  activityId and V2 column is named activityName. These columns are further used to column bind rows into the final data set.
```{r eval=FALSE}
names(X_test_train) <- sub_features$V2

# Rename columns Y_test_train to meaningful/descriptive names
Y_test_train <- rename(Y_test_train, activity = V1)

# Rename subject_test column to give descriptive name
subject_test_train <- rename(subject_test_train, subjectId = V1)
activity_labels <- rename(activity_labels, activityId = V1, activityName = V2)
```

I also use the merge() function to join the activity column of Y_test_train and the activityID column of activity_labels so that I can get the activityName column in the Y_test_train data set.
```{r eval=FALSE}
# Merge activity labels into Y_test_train
Y_test_train <- merge(Y_test_train, activity_labels, by.x = "activity", by.y = "activityId", all = TRUE)
```

#### 4. Combine all the dataset to arrive at the final data set.
Now that the columns/variables have been renamed, I arrive at the final data set by just combining the columns of the individual data sets using the cbind() function. I also use the select() function to remove the activity variable(containing ids or numbers) from the final data set "final_data" so that we do not have both activity names and the ids corresponding to them in the same data set(they represent the same data, so two columns are not required)
```{r eval=FALSE}
# Combine all the data by column binding the obtained data tables
final_data <- cbind(subject_test_train, Y_test_train, X_test_train)
final_data <- select(final_data, -activity)
```

#### 5. Appropriately label the data set with descriptive variable names.
For showing descriptive names in the final data set, I use a series of gsub commands which do the following.

1. 't' at the start of a variable is change to 'Time'
2. 'f' at the start of a variable is change to 'Frequency'
3. '-', '(' and ')' symbols are removed from the variable names
4. 'mean' is changed to 'Mean'
5. 'std' is changed to 'Standard'
6. 'Acc' is changed to 'Accelerometer'
7. 'Gyro' is changed to 'Gyroscope'
8. 'Mag' is changed to 'Magnitude'
9.  BodyBody' is changed to 'Body'


Example renaming is as below:
```{r eval=FALSE}
names(final_data)<-gsub("-", "", names(final_data))
names(final_data)<-gsub("\\(", "", names(final_data))
names(final_data)<-gsub(")", "", names(final_data))
names(final_data)<-gsub("mean", "Mean", names(final_data))
```

I also change the X, Y and Z at the end of the variable names to 'InXaxis', 'InYaxis' and 'InZaxis' to make it more meaningful.

```{r eval=FALSE}
names(final_data)<-gsub("X$", "InXaxis", names(final_data))
```

#### 6. Write the data set final_data to file
I use the following commands to write the final_data data set to file

```{r eval=FALSE}
setwd("./")
write.table(final_data, file = "First_Tidy_Dataset_Unsummarized.txt", row.names = FALSE)
```

#### 7. Create a second, independent tidy data set with the average of each variable for each activity and each subject.
Now, I use the group_by and summarize_each functions to group the data set final_data on the basis of activity and subjects and then find the average of the measurements.

```{r eval=FALSE}
final_data_summary <- (final_data %>% 
                         group_by(activityName, subjectId) %>% 
                         summarise_each(funs(mean))
                      )
```

At the end, write the final summarized data set to file
```{r eval=FALSE}
# Write the second data set to file
setwd("./")
write.table(final_data_summary, file = "Second_Independent_Summarized_Dataset.txt", row.names = FALSE)
```


